
---
title: "Multiple Regression Analysis in R"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships, such as how different factors collectively impact outcomes like election results, economic indicators, or sports performances. By understanding these relationships, journalists can provide deeper insights and more accurate predictions.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Data Import

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains various statistics for NFL teams from the 2019 season. This dataset will be used to build our model.

```{r}
teams <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/main/nfl_2019.csv")
```

## Model Building

Next, we construct a linear model to analyze how different factors like `Yards Gained`, `Takeaways`, and `Points Scored` influence the number of wins. This model helps us quantify the relationships and assess which factors are significant predictors of success.

```{r}
model <- lm(Wins ~ `Yards Gained` + Takeaways + `Points Scored`, data = teams)
summary(model)
```

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For your model:

*   **Min and Max** values indicate the range of residuals. A maximum of 4.1361 and a minimum of -4.3780 suggest that there might be some substantial errors in predictions for some cases, but generally, the errors are relatively evenly distributed around zero.
*   **Median** close to zero (-0.0604) indicates that the model is generally centered around the actual data points, but still, some deviations exist.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

*   **Yards Gained**: The coefficient is -0.001329. This suggests a very small negative impact of yards gained on wins, but this effect is not statistically significant (p = 0.34332), implying that as yards gained increase, the number of wins slightly decreases, which is counterintuitive and likely indicates no real effect or that the model isn't capturing the complexities.
*   **Takeaways**: The coefficient is 0.150593, suggesting that each additional takeaway is associated with an increase in wins by about 0.151, which is near the threshold of significance (p = 0.07928). This suggests a potentially meaningful, though not conventionally statistically significant, positive effect on wins.
*   **Points Scored**: The coefficient is 0.038732, which is significant (p = 0.00665). This indicates that each additional point scored is associated with an increase in wins by about 0.039, suggesting a clear and meaningful impact.

### Model Fit (3 points)

*   **Residual Standard Error**: The RSE of 2.072 tells us that the typical prediction by the model deviates from the actual number of wins by about 2.072 wins.
*   **Multiple R-squared**: 0.6232 indicates that about 62.32% of the variability in wins is explained by the model, which is relatively high, suggesting a good fit.
*   **Adjusted R-squared**: 0.5828 adjusts the R-squared for the number of predictors and indicates that about 58.28% of the variance is explained by the model after accounting for the number of variables used.
*   **F-statistic**: The F-statistic of 15.44 with a very low p-value (4.058e-06) suggests the model as a whole is statistically significant.

Is this a useful model, based on what you've seen? Why?

This model is somewhat useful, but it definitely has some limitations that make it feel incomplete. The R-squared is about 0.62, which means it has a good amount of the variation in wins, and the model overall is statistically significant, which suggests that at least some of the variables are meaningfully related to team success. Points scored stands out the most, since it is statistically significant and aligns with what should be expected: teams that score more tend to win more. Yards gained has a negative and insignificant coefficient, which doesn’t really make sense and suggests the model isn’t fully capturing what actually leads to wins. Takeaways seem to have a positive effect, but they are not quite statistically significant, so it’s harder to interpret that relationship. On top of that, the average error of about two wins shows that the predictions are not fully accurate. Overall, the model does give helpful insight, especially about scoring, but it should be interpreted carefully and would need more work before being used to make strong claims.

## Correlation Analysis

After building the model, we calculate the correlation matrix for our variables of interest. This matrix helps visualize the strength and direction of the relationships between each pair of variables.

```{r}
selected_vars <- teams |>  select(`Yards Gained`, Takeaways, `Points Scored`, Wins)
cor(selected_vars)
```

## Visualization of Correlation Matrix

Finally, we use a visualization tool to plot the correlation matrix. This graphical representation makes it easier to interpret the correlations at a glance, showing which variables are most closely connected.

```{r}
corr_matrix <- cor(selected_vars)
corrplot(corr_matrix, method = "circle")
```

## Results Interpretation (4 points)

The results from our correlation analysis show the strength of the relationships between the number of wins and factors like yards gained, takeaways, and points scored. High positive correlations indicate strong relationships where increases in one variable are associated with increases in another. Conversely, negative correlations suggest an inverse relationship. By understanding these dynamics, journalists can better explain the factors that contribute to a team's success or failure in a season.

Which of these three factors - Yards Gained, Takeaways and Points Scored - have the most impact on wins? Are there any that do not seem impactful at all?
 
Based on the correlation results, points scored clearly has the strongest impact on wins, with a correlation of about 0.71. This indicates a strong positive relationship, meaning teams that score more points tend to win more games, which makes sense and matches what we saw in the regression model. Takeaways also show a meaningful relationship with wins, with a moderate positive correlation of around 0.57, suggesting that teams that force more turnovers generally perform better, even if this effect was slightly weaker in the regression.
On the other hand, yards gained appears to be the least impactful out of the three. Its correlation with wins is only about 0.44, which is noticeably weaker than the others. While this still shows some positive relationship, it is not as strong, and combined with its poor performance in the regression model, it suggests that yards gained alone does not do a great job of explaining wins. To sum it up, points scored has the most impact, takeaways have a moderate impact, and yards gained seems the least meaningful in this analysis.

## Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this model to cover a team or game


This analysis could be used to explain why a team wins or loses beyond the final score. Through comparing points scored, takeaways, and yards gained, the factors that actually drove the outcome are highlighted, such as showing that a team won because they scored efficiently or forced turnovers, even if the yardage was similar. Over the course of a season, it can also help identify patterns, like teams that gain a lot of yards but don’t turn that into points, adding clearer context to your coverage.