---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```




### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```

### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |>
  mutate(pct_change = (`Fall 2024` - `Fall 2021`) / `Fall 2021` * 100)
 
print(umd_fees_pct)
umd_fees_pct25 <- umd_fees |>
  mutate(pct_change = (`Fall 2025` - `Fall 2024`) / `Fall 2024` * 100)
 
```



### Task 4: Describe our results


Write a few sentences about what you found in the data. What's the story here?

 
I notice that after 2021 is when the student counseling center received fees beginning after 2021, which makes the percent change unable to be read. The differences between 2024 and 2025 are not that great, though the difference in yearly changes are stark. The shuttle bus and sustainability center overall experienced the most change in fees. I think after more awareness on sustainability over the last few years, this topic makes a lot of sense. The shuttle bus fees probably have a lot to do with mobility for students on and around campus. 





```{r}
comp <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/class_composition.csv")

comp |>
  summarize(
    mean_first = mean(first_year),
    mean_soph = mean(sophomore),
    mean_junior = mean(junior),
    mean_senior = mean(senior),
    mean_grad = mean(grad)
  )
```
```{r}
old_value <- 10
new_value <- 7

percent_change <- ((new_value - old_value) / old_value) * 100
print(percent_change)
```
```{r}
old_value <- 8
new_value <- 9

percent_change <- ((new_value - old_value) / old_value) * 100
print(percent_change)
```
```{r}
old_value <- 7
new_value <- 7

percent_change <- ((new_value - old_value) / old_value) * 100
print(percent_change)
```
```{r}
old_value <- 9
new_value <- 9.5

percent_change <- ((new_value - old_value) / old_value) * 100
print(percent_change)
```

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
