---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Holly Pommett"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}

inspections <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- inspections %>%
  summarise(
    mean = mean(compliance_score, na.rm = TRUE),
    sd = sd(compliance_score, na.rm = TRUE),
    min = min(compliance_score, na.rm = TRUE),
    max = max(compliance_score, na.rm = TRUE)
  )
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
mean_score <- mean(inspections$compliance_score, na.rm = TRUE)

ggplot(inspections, aes(x = compliance_score)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  geom_vline(xintercept = mean_score, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Restaurant Compliance Scores",
    subtitle = paste("Mean score:", round(mean_score, 2)),
    x = "Compliance Score",
    y = "Number of Inspections"
  )
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**
The histogram shows us a left-skewed data set, with most restaurants having high compliance scores with health inspections. The standard deviation here is high, at a score of 5.83, which makes me wonder about the restaurants on the lower tail side of compliance. A next step in reporting could be an investigative piece, exploring the failures of these restaurants with compliance of safety precautions to the general public, perhaps compared to their popularity and survival in the restaurant market with quotes on the reality of working there from ex-employees. 
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily %>%
  summarize(
    mean_bus = mean(bus, na.rm = TRUE),
    sd_bus = sd(bus, na.rm = TRUE),
    mean_rail = mean(rail, na.rm = TRUE),
    sd_rail = sd(rail, na.rm = TRUE)
  )

wmata_stats
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily %>%
  mutate(
    rail_zscore = (rail - mean(rail, na.rm = TRUE)) / sd(rail, na.rm = TRUE),
    bus_zscore = (bus - mean(bus, na.rm = TRUE)) / sd(bus, na.rm = TRUE)
  )

wmata_zscores %>% 
  select(date, rail, rail_zscore, bus, bus_zscore) %>% 
  arrange(desc(abs(rail_zscore)))
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)
There are 6 unusual days.

```{r}
# Check great than 2 
unusual_rail_days <- wmata_zscores %>%
  filter(abs(rail_zscore) > 2) %>%
  arrange(desc(abs(rail_zscore)))

# Results
unusual_rail_days

# Unusual
nrow(unusual_rail_days)
```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(42) # Using 42 as common seed

wmata_sample <- wmata_daily %>%
  slice_sample(n = 91)

# Sample statistics ?
sample_stats <- wmata_sample %>%
  summarize(
    sample_mean_bus = mean(bus, na.rm = TRUE),
    sample_sd_bus = sd(bus, na.rm = TRUE),
    sample_mean_rail = mean(rail, na.rm = TRUE),
    sample_sd_rail = sd(rail, na.rm = TRUE)
  )

sample_stats
```
The sample mean and standard deviation should be pretty close to the population statistics, though they will not be identical based on the randomization of the two samples. This result suggests random sampling is a good tool for estimation when there is only a smaller portion of data available, but not completely representative of statistical trends.
### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)
The most significant day of unusual z-scores that I noticed in my data was on January 1st, 2024, as well as January 1st, 2025. This day had the most significant difference for both modes of transportation, coming in around -2.4 for rails in 2024, -2.12 for buses in 2024, -2.17 for rails in 2025, and -1.78 for buses in 2025. A good story lead could be diving deeper into the fact that this was New Year's Day; this deviation is significant, perhaps suggesting that the holiday threw people off work and school schedules and consequently affected the transit ridership.

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}

md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}

md_car_thefts <- md_car_thefts %>%
  rename(theft_count = 1) %>% 
  mutate(theft_rate = (as.numeric(theft_count) / as.numeric(population)) * 10000)
```

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}

# statewide summary
summary_stats <- md_car_thefts %>% 
  summarize(
    median_rate = median(theft_rate, na.rm = TRUE),
    total_thefts = sum(as.numeric(theft_count), na.rm = TRUE) 
  )


summary_stats

# counties above median rate
md_car_thefts %>% 
  filter(theft_rate > median(theft_rate, na.rm = TRUE)) %>% 
  arrange(desc(theft_rate))
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)
(Unfortunately I think I did this wrong I am not sure if I located the correct data and ran into some problems with the code here and having to rename things)
 
 According to the analysis on the environment, the distribution of the car theft population tends to run more in smaller, urban centers. The data might be skewed by larger outliers like the city, or smaller outliers like rural areas. A good lede would be to investigate the concentration of auto theft in urban areas, and see if there are any consistencies in the types of vehicles stolen or the motive behind it. 

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:


### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

Given the mean and standard deviation, I would prioritize identifying any outliers and calculating the range of this dataset. Since the editor suspects poor business within the community pools, I would want to know if the distribution is skewed or if only select pools have larger standard deviations below the calculated mean. I would also look at the median to determine if the average of 187 is being inflated by more popular pools, narrowing the search to which pools may be underperforming overall.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)
A pool with a z-score of −2.0 is performing significantly worse than the average pool in this sample. This turns out to be more useful than a simple ranking because that will not reveal reality of the gaps between businesses. A pool could be ranked last but still be within a normal range of business in comparison to the others. The z-score would be useful in determining real issues versus a less popular couple of days in the randomization of the 60-day sample. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would first investigate the sampling timeframe; consider the first 60 days of summer might be prime vacation time, away from local businesses, or maybe instead that is when the use of community pools skyrocket due to the excitement of visiting for the first time this year. Perhaps weather patterns affected the select days, with particular thunderstorms canceling pool usage, or extreme heatwaves causing the concentration of customers at certain pools. The margin of error would be my next concern, given that sampling errors do occur when the entire timeline is not accounted for. I would definitely consider all these factors in my story, wanting to represent the most accurate depiction of why community pools might be underperforming this summer. 

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
