---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE: The data set contains 404 records of White House employees with five variables: employee name, employment status, position title, pay basis, and salary. From a reporter standpoint, what stands out is the range of salaries and the presence of salaries listed as 0, which likely represent unpaid roles, like volunteer or nonprofit, or positions not receiving a traditional salary. This void in data may disrupt calculations like the mean and should be considered carefully during analysis.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries %>%
  summarize(
    mean_salary = mean(SALARY, na.rm = TRUE),
    median_salary = median(SALARY, na.rm = TRUE)
  )
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries %>%
  summarize(
    mean_salary   = mean(SALARY, na.rm = TRUE),
    median_salary = median(SALARY, na.rm = TRUE),
    min_salary    = min(SALARY, na.rm = TRUE),
    max_salary    = max(SALARY, na.rm = TRUE)
  )
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE The mean salary is higher than the median salary, which suggests the distribution is skewed by a much smaller number of high income earners. The minimum salary is $0, indicating that some White House staffers are unpaid or serve in volunteer or nonprofit roles. Meanwhile, the maximum salary is significantly higher than most other salaries in the data set. With the sharp contrast of salary listings, there is potential of distorted summary statistics, particularly the mean.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE A compelling story could focus on the salary disparities within the White House, highlighting how a much smaller group of officials earn far more in comparison to most employees. The next step would be to analyze which positions or offices command the highest pay and explore differences in compensation by job role, employment status and pay structure. Additional context could be provided by comparing these salaries to those from previous administrations or by interviewing a current or former staff member for a more personal interaction. 
