In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
# Calculate mean and standard deviation
mean_danceability <- mean(songs$danceability, na.rm = TRUE)
sd_danceability <- sd(songs$danceability, na.rm = TRUE)

# Display results
mean_danceability
sd_danceability
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
ggplot(songs, aes(x = danceability)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "white") +
  labs(
    title = "Distribution of Danceability Scores",
    x = "Danceability",
    y = "Count"
  ) +
  theme_minimal()
```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
ggplot(songs, aes(x = danceability)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "white") +
  geom_vline(xintercept = mean_danceability, 
             color = "red", 
             size = 1) +
  labs(
    title = "Distribution of Danceability Scores",
    x = "Danceability",
    y = "Count"
  ) +
  theme_minimal()
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
ggplot(songs, aes(x = danceability)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "white") +
  geom_vline(xintercept = mean_danceability, 
             color = "red", 
             size = 1) +
  geom_vline(xintercept = mean_danceability + sd_danceability,
             color = "green",
             linetype = "solid",
             size = 1) +
  geom_vline(xintercept = mean_danceability - sd_danceability,
             color = "green",
             linetype = "solid",
             size = 1) +
  labs(
    title = "Distribution of Danceability Scores",
    x = "Danceability",
    y = "Count"
  ) +
  theme_minimal()
```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

The danceability levels cluster around the middle to higher range, rather than forming a normal distribution. The distribution appears to be skewed toward more danceable songs, meaning many songs in the dataset are relatively upbeat or energetic. The mean represents the average song’s danceability, and most songs fall within one standard deviation of that average, showing the majority share similar energy and dance-friendly qualities.

In 2020, popular music leaned heavily toward danceable, high-energy tracks. Even during the unprecedented pandemic during that year, listeners appeared to favor songs that felt fun and mood-lightening. This data could reflect the public's turn to upbeat music as a source of escape during hard times.

### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

