---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)


```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
names(mandms)
```

### Task 3: Sample 5 rows of data from the dataset

```{r}
sample_mandms <- mandms %>%
  sample_n(5)

sample_mandms

```
### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms
```{r}
mean(sample_mandms$black)
sd(sample_mandms$black)
```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
mean(mandms$black)
sd(mandms$black)
```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The sample mean and standard deviation are similar to the population mean and standard deviation, but they are not exactly the same. The sample only includes five observations, while the population includes all observations in the data set. Small random samples can vary slightly from the population values.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The analysis shows that the mean and standard deviation of the sample are close to the population values, but not identical. This demonstrates how random sampling works in statistics. A small sample can provide an estimate of the population, but the results aren't totally accurate based on the randomization. Larger samples would likely produce statistics that are even closer to the true population values.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
