

---
title: "final.project.2"
output: html_document
---

---
title: "final_project"
output: html_document
---

# Final Project, Data Setup and Exploration

## Introduction
---
title: "UMD Carbon Emissions and Offset Dependence"
output: html_document
---

# Introduction

This project examines greenhouse gas emissions trends at the University of Maryland from 2005 to 2025. The goal is to analyze how emissions have changed over time and evaluate whether reported progress toward carbon neutrality reflects actual emissions reductions or increasing reliance on accounting mechanisms such as renewable energy credits and offsets.

# Hypothesis

UMD's reported progress toward carbon neutrality increasingly relies on purchased offsets and accounting adjustments rather than direct emissions reductions, meaning the gap between gross emissions and net reported emissions has grown over time.

# Data Source

The dataset is based on the University of Maryland Climate Action Plan dashboard:
https://sustainingprogress.umd.edu/ClimateActionPlan

Because the dashboard does not provide a downloadable full CSV, the data was manually reconstructed from reported annual emissions summaries. Values are consistent with published trends but should be interpreted as approximations for analytical purposes.

This dashboard provides annual greenhouse gas emissions data categorized by Scope 1, Scope 2, and Scope 3 emissions. 
Scope 1: On-campus emissions 
Scope 2: Purchased electricity 
Scope 3: Indirect emissions The data spans from 2005 (baseline year) through 2025.

# Load libraries

```{r}


library(tidyverse)

umd <- tribble(
  ~Year, ~Scope1, ~Scope2, ~Scope3, ~Total,
  2005, 180000, 120000, 20000, 320000,
  2006, 175000, 115000, 21000, 311000,
  2007, 170000, 110000, 22000, 302000, 
  2008, 179000,121000,21500,321500, 
  2009, 177000,120000,22000,319000, 
  2010,175000,118000,22500,315500,
  2011,173000,116000,23000,312000,
  2012,170000,114000,23500,307500,
  2013,168000,112000,24000,304000,
  2014,165000,110000,24500,299500,
  2015,162000,108000,25000,295000,
  2016,160000,105000,25500,290500,
  2017,158000,100000,26000,284000,
  2018,155000,95000,26500,276500,
  2019,152000,90000,27000,269000,
  2020,150000,0,26000,176000,
  2021,147000,0,25500,172500,
  2022,145000,0,25000,170000,
  2023,142000,0,24500,166500,
  2024,140000,0,24000,164000,
  2025,138000,0,23500,161500
)



write_csv(umd, "umd_emissions.csv")
```



#Chart 

```{r}
read_csv("umd_emissions.csv")
```
```{r}
umd <- umd %>%
  mutate(
    gross_emissions = Scope1 + Scope2 + Scope3,

    # Structural electricity shift after 2020
    electricity_shift = lag(Scope2, default = first(Scope2)) - Scope2,
    electricity_shift = ifelse(electricity_shift < 0, 0, electricity_shift),

    # Offsets
    estimated_offsets = case_when(
      Year >= 2020 ~ electricity_shift * 0.8,
      TRUE ~ 0
    ),

    # Net emissions after adjustments
    net_emissions = gross_emissions - estimated_offsets,

    # Hypothesis testing
    gap = gross_emissions - net_emissions,
    offset_ratio = estimated_offsets / gross_emissions,

    baseline = first(gross_emissions),
    percent_reduction = (1 - net_emissions / baseline) * 100
  )


#Graph/Plot

umd %>%
  pivot_longer(cols = Scope1:Scope3,
               names_to = "Scope",
               values_to = "Emissions") %>%
  ggplot(aes(x = Year, y = Emissions, color = Scope)) +
  geom_line()
```

 This dataset is reconstructed based on publicly available UMD Climate Action Plan summaries. Exact annual values are not fully published in a single downloadable dataset, so values I used approximate values and reported long-term trends in Scope 1, Scope 2, and Scope 3 emissions. The data trends are still consistent with the fully published official version. 
 
 The data set covers three scopes of UMD's plan to reduce carbon emissions. Scope 1 covers emissions from on-campus fossil fuel use like heating, fleet, and campus operations. Scope 2 shows numbers from the emissions of purchased electricity, which is especially important to this story because it drops to near zero after the shift to renewable energy and shows quantifiable change. Scope 3 is other indirect emissions such as commuting, travel, and waste, but is harder to fully measure and is primarily based off estimates from the school. Overall, I think for this project I will be focused on Scopes 1 and 2, given the fact that they are the most reliable and show the most impactful change. Scope 3 is also useful but less central, since it is more variable and harder to directly control through campus policy.
 
 
Some issues with this data set is the lack of distinguishing between offset and actual reduction emissions, which is a big part of actually seeing the carbon measurement truly shrink. This part of carbon neutrality is definitely a topic that would have to be explained in a story to the general public. Also, Scope 3 may be inconsistently measured over the years, given the change in policies and different methods of measurements. 

I think my biggest question that remains is investigating how much of the reported carbon neutrality is achieved through actual emissions reductions vs. carbon offsets. This is critical to my reporting for this project, and shows the difference between reducing carbon emissions, or just "making them up" through environmental efforts. 


```{r}
head(umd)
```

```{r}
ggplot(umd, aes(x = Year)) +
  geom_line(aes(y = Scope1, color = "Scope 1")) +
  geom_line(aes(y = Scope2, color = "Scope 2")) +
  geom_line(aes(y = Scope3, color = "Scope 3")) +
  labs(title = "UMD Emissions by Scope Over Time",
       y = "Emissions",
       color = "Scope")

umd %>%
  pivot_longer(cols = c(gross_emissions, net_emissions),
               names_to = "Type",
               values_to = "Emissions") %>%
  ggplot(aes(x = Year, y = Emissions, color = Type)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Gross vs Net Emissions Over Time",
    x = "Year",
    y = "Emissions"
  )
```
This result here is really important because I can see the equal values between the lines until the change in 2020, until it rebalances itself. Net emissions really are a huge focus in this project, as I can tell that the overall amount of emissions from UMD has sharply decreased since the change in Scope 2, showing actual results in the reduction of UMD carbon emissions and maintenance of carbon neutrality. 

```{r}
umd <- umd %>%
  mutate(gross_emissions = Scope1 + Scope2 + Scope3)

ggplot(umd, aes(x = Year, y = gross_emissions)) +
  geom_line() +
  geom_point() +
  labs(title = "Gross Emissions Over Time")
```
This result here shows the sharp decline in gross emissions over time. This will be helpful in my conclusion to demonstrate the actual impact policy change had on the results in 2020. 

```{r}
model_gross <- (gross_emissions ~ Year, data = umd)
summary(model_gross)
```

```{r}
model_offset <- (offset_ratio ~ Year, data = umd)
summary(model_offset)
```

After doing more research I need to adjust the last two codes. The other graphs we've covered similarly in class, but the gross summary and offset values I need to understand better. 

Overall, I need to also better understand the amount of purchases UMD must make to offset this balance and maintain neutrality. For such a large campus, I know that they do rely heavily on offsets.
